\textbf{Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated}

### 2023-04-03 analysis.Rmd
Add check against norms

```{r message=FALSE, warning=FALSE}
outliers <- c(
  # More than 2 attention check fails
  "611d03b822d4c8e041ea0c32_m0knb",
  "5f0f0a2a8b2a480447f31b21_lqgpz",
  # Very short duration for questionnaire + low rating correlations (< 0.1)
  "5eaef8702b68455d6e130595_ptsga",
  "61356174c090b0083f131e01_asm81",
  # Prefered not answering to sexual orientation: further analysis impossible
  "60eb34f117838a34a29a69d3_rxv85"
)
outliers_partial <- c(
  # 1 attention check failed
  "5dc3485219ca0326027ce91f_37ho9",
  "5c6414540821d30001046198_x9q7r",
  "60dd7b03f1e72d38230df476_9yh9n",
  "5962799cb752840001ca478b_jh4sl",
  "5f44c23fbf2ddb80bcdf0edc_dnbny",
  "5e80370d48b5f47170e30e5c_5w2gf"
)
```

@@

  # Plot
  p <- df |>
    ggplot(aes_string(x = var, y = "Real")) +
    stat_density_2d(aes(fill = ..density..), geom = "raster", contour = FALSE) +
    scale_fill_gradientn(colors = c("white", fill), guide = "none") +
    ggnewscale::new_scale_fill() +
    geom_hline(yintercept = 0.5, linetype = "dotted") +
    # geom_point2(alpha = 0.25, size = 4, color = "black") +
    geom_line(data = y_conf, aes(y = Predicted, group = interaction(Belief, Sex), color = Sex), linetype = "dashed") +
    geom_ribbon(data = y_real, aes(y = Predicted, group = Sex, fill = Sex, ymin = CI_low, ymax = CI_high), alpha = 1 / 3) +
    geom_line(data = y_real, aes(y = Predicted, color = Sex), size=1) +
    # geom_ribbon(data = trend, aes(y = Predicted, group=Sex, fill=Sex, ymin = CI_low, ymax = CI_high), alpha = 1/6) +
    # geom_line(data = trend, aes(y = Predicted, color=Sex, linetype=Trend, group=interaction(Sex, group)), size=0.6) +
    geom_text(data = sig1, aes(y = y, x = x, label = label, color = Sex), size = ifelse(sig1$p < .05, 8.5, 3.5)) +
    geom_text(data = sig2, aes(y = y, x = x, label = label, color = Sex), size = ifelse(sig2$p < .05, 5, 3)) +
    scale_color_manual(values = c("Male" = "#2196F3", "Female" = "#E91E63")) +
    scale_fill_manual(values = c("Male" = "#2196F3", "Female" = "#E91E63")) +
    scale_x_continuous(expand = c(0, 0), labels = scales::percent) +
    scale_y_continuous(expand = c(0, 0), breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c("Fake", "25%", "50%", "75%", "Real")) +
    labs(y = "Simulation Monitoring") +
    guides(fill = guide_legend(override.aes = list(alpha = 1))) +
    theme_modern(axis.title.space = 5) +
    ggside::geom_xsidedensity(fill = fill, color = "white") +
    ggside::geom_ysidedensity(fill = "#9C27B0", color = "white") +
    ggside::theme_ggside_void() +
    ggside::scale_ysidex_continuous(expand = c(0, 0)) +
    ggside::scale_xsidey_continuous(expand = c(0, 0))

@@

>>>>

```{r message=FALSE, warning=FALSE}
outliers <- c(
  # More than 2 attention check fails
  "S113",
  "S049",
  # Very short duration for questionnaire + low rating correlations (< 0.1)
  "S035",
  "S117",
  # Prefered not answering to sexual orientation: further analysis impossible
  "S101"
)
outliers_partial <- c(
  # 1 attention check failed
  "S027",
  "S017",
  "S097",
  "S006",
  "S055",
  "S033"
)
```

@@

  # Plot
  p <- df |>
    ggplot(aes(x = .data[[var]], y = .data[["Real"]])) +
    stat_density_2d(aes(fill = after_stat(density)), geom = "raster", contour = FALSE) +
    scale_fill_gradientn(colors = c("white", fill), guide = "none") +
    ggnewscale::new_scale_fill() +
    geom_hline(yintercept = 0.5, linetype = "dotted") +
    # geom_point2(alpha = 0.25, size = 4, color = "black") +
    geom_line(data = y_conf, aes(y = Predicted, group = interaction(Belief, Sex), color = Sex), linetype = "dashed") +
    geom_ribbon(data = y_real, aes(y = Predicted, group = Sex, fill = Sex, ymin = CI_low, ymax = CI_high), alpha = 1 / 3) +
    geom_line(data = y_real, aes(y = Predicted, color = Sex), linewidth=1) +
    # geom_ribbon(data = trend, aes(y = Predicted, group=Sex, fill=Sex, ymin = CI_low, ymax = CI_high), alpha = 1/6) +
    # geom_line(data = trend, aes(y = Predicted, color=Sex, linetype=Trend, group=interaction(Sex, group)), size=0.6) +
    geom_text(data = sig1, aes(y = y, x = x, label = label, color = Sex), size = ifelse(sig1$p < .05, 8.5, 3.5)) +
    geom_text(data = sig2, aes(y = y, x = x, label = label, color = Sex), size = ifelse(sig2$p < .05, 5, 3)) +
    scale_color_manual(values = c("Male" = "#2196F3", "Female" = "#E91E63")) +
    scale_fill_manual(values = c("Male" = "#2196F3", "Female" = "#E91E63")) +
    scale_x_continuous(expand = c(0, 0), labels = scales::percent) +
    scale_y_continuous(expand = c(0, 0), breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c("Fake", "25%", "50%", "75%", "Real")) +
    labs(y = "Simulation Monitoring") +
    guides(fill = guide_legend(override.aes = list(alpha = 1))) +
    theme_modern(axis.title.space = 5) +
    ggside::geom_xsidedensity(aes(fill = Sex), alpha=2/3, color = NA) +
    ggside::geom_ysidedensity(aes(fill = Sex), alpha=2/3, color = NA) +
    ggside::theme_ggside_void() +
    ggside::scale_ysidex_continuous(expand = c(0, 0)) +
    ggside::scale_xsidey_continuous(expand = c(0, 0))

@@

### Cross-Validate with Normmative Scores

```{r make_models_norms, message=FALSE, warning=FALSE, fig.width=fig.width, fig.height=fig.width, cache=cache}
rez_at_norm <- filter(df, Stimulus_Interest == TRUE) |> 
  mutate(Norms_Attractive = normalize(Norms_Attractive)) |> 
  make_model(
    formula = "Sex / poly(Norms_Attractive, 2) + (1|Participant) + (1|Stimulus)",
    var = "Norms_Attractive", fill = "#F44336"
)

rez_at_norm$p

rez_tr_norm <- filter(df, Stimulus_Interest == TRUE) |> 
  mutate(Norms_Trustworthy = normalize(Norms_Trustworthy)) |> 
  make_model(
    formula = "Sex / poly(Norms_Trustworthy, 2) + (1|Participant) + (1|Stimulus)",
    var = "Norms_Trustworthy", fill = "#4CAF50"
)

rez_tr_norm$p
```

### 2023-04-03 submission/manuscript.Rmd
Add check against norms

keywords          : "attractiveness, simulation monitoring, fiction, deep fakes, sense of reality"
wordcount         : "4088"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

While not all simulations have achieved perfect realism [e.g., Computer Generated Images - CGI in movies often lack certain key details that makes them visually distinct from real images, @mcdonnell2010face], it is fair to assume that these technical limitations will become negligible in the near future, particularly in the field of face generation [@nightingale2022; @tucciarelli2020; @moshel2022]. Such performance, however, leads to a new question: if real and fake stimuli cannot be distinguished based on their objective characteristics, how can we make judgements regarding their nature?

@@

Literature shows that the context surrounding a stimulus often plays an important role in the assessment of its reality [a process henceforth referred to as *simulation monitoring*, @makowski2018cognitive; @makowski2019phenomenal]. With the extensive search and processing of cues within ambiguous stimuli being an increasingly complex and cognitively effortful strategy [@michael2021source; @susmann2021persuasion], people tend to draw on peripheral contextual cues (**Figure 1**), such as the source of the stimulus (e.g., in what journal has information been published), and its credibility, authority and expertise, to help facilitate their evaluation [@petty1986elaboration; @susmann2021persuasion; @michael2021source]. However, the automization and decontextualization of information allowed by online social media (where text snippets or video excerpts are mass-shared with little context) makes this task increasingly difficult [@berghel2018weaponizing; @chen2015news]. Thus, in the absence of clear contextual information, what drives our beliefs of reality?

@@

Beyond stimulus- and individual-related characteristics, evidence suggests that the interaction between the two (i.e., the subjective reaction associated with the experience of a given stimulus), contributes to simulation monitoring decisions. For instance, the intensity of experienced emotions have been shown to increase one's sense of presence - the extent to which one feels like "being there", as if the object of experience was real - when engaged in a fictional movie or a VR environment [@sanchez2005presence; @makowski2017being]. Indeed, participants' self-reported emotional arousal were found to significantly predict the probability that they would perceive images as real [@azevedo2020body]. Conversely, beliefs that emotional stimuli were fake (e.g., that emotional scenes were not authentic but instead involved actors and movie makeup) were found to result in emotion down-regulation [@makowski2019phenomenal; @sperduti2017distinctive]. In line with these findings, studies on susceptibility to fake news have also found heightened stimulus emotionality to be associated with greater belief [@martel2020reliance; @bago2022emotion]. In fact, photojournalistic images labelled fake were associated with greater surprise when they were highly arousing [@azevedo2020body]. Additionally, other factors, such as the stimuli's perceived self-relevance [@sperduti2016paradox; @goldstein2009pleasure], as well as familiarity [@begg1992dissociation], could also play a role in guiding our appraisal of a stimulus.

@@

AI-generated images of faces, due to their popularity as a target of CGI technology and to the possibility of experimentally manipulating facial features, are increasingly used to study face processing in relationship with saliency or emotions, as well as to other important components of face evaluation, such as trustworthiness or attractiveness [@sobieraj2014beautiful; @balas2017; @tsikandilakis2019beauty; @calbi2017]. Interestingly, artificially created faces rated as more attractive (by an independent group of raters) were perceived as less real [@tucciarelli2020]. Conversely, @liefooghe2022faces reports that attractiveness ratings were significantly lower when participants who were told that the faces were AI-generated were compared to those who had no prior knowledge. Whereas this line of evidence suggests that reality beliefs have an effect on face attractiveness ratings, the opposite question - whether attractiveness could drive simulation monitoring - has received little attention to date.

@@

@@

<!-- In line with open-science standards, all the material (stimuli generation code, experiment code, raw data, analysis script with complementary figures and analyses, preregistration, etc.) is available at [**https://github.com/RealityBending/FakeFace**](https://github.com/RealityBending/FakeFace){.uri}. -->

@@

One hundred and fifty participants were recruited via *Prolific*, a crowd-sourcing platform recognized for providing high quality data [@peer2022]. The only inclusion criterion was a fluent proficiency in English to ensure that the experiment instructions would be well-understood. Participants were incentivised with a reward of about \textsterling 7.5 for completing the study, which took about 45 minutes to finish. Demographic variables (age, gender, sexual orientation, education and ethnicity) were self-reported on a voluntary basis.

We excluded 5 participants that either failed 2 (\>= 66%) or more attention check questions, took an implausibly short time to finish the questionnaires or had incomplete responses. The final sample included 145 participants (Mean age = 28.3, SD = 9.0, range: [19, 66]; Sex: 48.3% females, 51.0% males, 0.7% others).

@@

The real-fake ratings (measured originally on a [-1, 1] analog scale) were converted into two scores, corresponding to two conceptually distinct mechanisms: the dichotomous *belief* (real or fake, based on the sign of the rating) and the *confidence* (the rating's absolute value) associated with that belief. The former was analyzed using logistic mixed models, which modelled the probability of assigning a face to the real (\>= 0) as opposed to fake (\< 0). The latter, as well as the other face ratings (attractiveness, beauty, trustworthiness and familiarity), was modelled using mixed beta regressions (suited for outcome variables expressed in percentages). The models included the participants and stimuli as random factors.

@@

### Manipulation Check

@@

Several limitations have to be noted. The current experimental paradigm required participants to judge the realness of faces they had prior exposure to (which was done to prevent reality judgements from influencing the other ratings). Although the effect of re-exposure delay was negligible, the potential bias induced by face familiarity (as compared to judging completely new items) cannot be discarded. Future studies could examine that by incorporating novel face images or increasing the duration of the re-exposure delay. Moreover, the magnitude of the effects found in the study was relatively small, suggesting that the facial features measured in the study were not the key determinants of simulation monitoring. Hence, beyond exploring new potential mechanisms, future studies should include a more thorough debriefing to try to capture what conscious strategies (if any) the participants used (e.g., focusing on some features of the stimulus - like hair or eyes in the case of faces) to guide their reality beliefs.

>>>>

keywords          : "attractiveness, AI-generated images, fiction, ffafke news, sense of reality"
wordcount         : "4088"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

While not all simulations have achieved perfect realism [e.g., Computer Generated Images - CGI in movies or via recent algorithms such as GANs or diffusion models often include distortions or lack certain key details that makes them visually distinct from real images, @mcdonnell2010face; @corvi2022detection], it is fair to assume that these technical limitations will become negligible in the near future. This is particularly true in the field of face generation, where face-generation algorithms are already able to create stimuli that are virtually indistinguishable from real photos [@nightingale2022; @tucciarelli2020; @moshel2022]. Such technological feat, however, leads to a new question: if real and fake stimuli cannot be differentiated based on their objective "physical" characteristics, how can we form judgements regarding their nature?

@@

Literature shows that the context surrounding a stimulus often plays an important role in the assessment of its reality [a process henceforth referred to as *simulation monitoring*, @makowski2018cognitive; @makowski2019phenomenal]. With the extensive search and processing of cues within ambiguous stimuli being an increasingly complex and cognitively effortful strategy [@michael2021source; @susmann2021persuasion], people tend to draw on peripheral contextual cues (**Figure 1**), such as the source of the stimulus (e.g., in what journal has information been published), and its credibility, authority and expertise, to help facilitate their evaluation [@petty1986elaboration; @susmann2021persuasion; @michael2021source]. However, the atomization and decontextualization of information allowed by online social media (where text snippets or video excerpts are mass-shared with little context) makes this task increasingly difficult [@berghel2018weaponizing; @chen2015news]. Thus, in the absence of clear contextual information, what drives our beliefs of reality?

@@

Beyond stimulus- and individual-related characteristics, evidence suggests that the interaction between the two (i.e., the subjective reaction associated with the experience of a given stimulus), contributes to simulation monitoring decisions. For instance, the intensity of experienced emotions have been shown to increase one's sense of presence - the extent to which one feels like "being there", as if the object of experience was real - when engaged in a fictional movie or a VR environment [@sanchez2005presence; @makowski2017being]. Indeed, participants' self-reported emotional arousal were found to significantly predict the probability that they would perceive images as real [@azevedo2020body]. Conversely, beliefs that emotional stimuli were fake (e.g., that emotional scenes were not authentic but instead involved actors and movie makeup) were found to result in emotion down-regulation [@makowski2019phenomenal; @sperduti2017distinctive]. In line with these findings, studies on susceptibility to fake news have also found heightened stimulus emotionality to be associated with greater belief [@martel2020reliance; @bago2022emotion], and higher neurophysiological arousal was predictive of judging realistic images a real [@azevedo2020body]. Additionally, other factors, such as the stimuli's perceived self-relevance [@sperduti2016paradox; @goldstein2009pleasure], as well as familiarity [@begg1992dissociation], could also play a role in guiding our appraisal of a stimulus.

@@

AI-generated images of faces, due to their popularity as a target of CGI technology and to the possibility of experimentally manipulating facial features, are increasingly used to study face processing in relationship with saliency or emotions, as well as to other important components of face evaluation, such as trustworthiness or attractiveness [@sobieraj2014beautiful; @balas2017; @tsikandilakis2019beauty; @calbi2017]. Interestingly, artificially created faces rated as more attractive (by an independent group of raters) were perceived as less real [@tucciarelli2020]. Conversely, @liefooghe2022faces reports that attractiveness ratings were significantly lower when participants who were told that the faces were AI-generated were compared to those who had no prior knowledge. Whereas this line of evidence suggests that reality beliefs have an effect on face attractiveness ratings, the opposite question - whether attractiveness contributes to the formation of reality beliefs - has received little attention to date.

@@

All the material (preregistration, experiment demo, experiment code, raw data, analysis script with complementary figures and analyses, etc.) is available at [**https://github.com/RealityBending/FakeFace**](https://github.com/RealityBending/FakeFace){.uri}.

@@

@@

Although the main part of the study relied on within-subject design (with 109 trials per participant), we also planned to do between-participants analyses, thus aiming at collecting a larger sample than traditionally used in experimental psychology (with budget availability as the main constraint). One hundred and fifty participants were recruited via *Prolific*, a crowd-sourcing platform recognized for providing high quality data [@peer2022; douglas2023data]. The only inclusion criterion was a fluent proficiency in English to ensure that the experiment instructions would be well-understood. Participants were incentivised with a reward of about \textsterling 7.5 for completing the study, which took about 45 minutes to finish. Demographic variables (age, gender, sexual orientation, education and ethnicity) were self-reported on a voluntary basis.

We excluded 5 participants that either failed 2 (\>= 66.6%) or more attention check questions, took an implausibly short time to finish the questionnaires or had incomplete responses. The final sample included 145 participants (Mean age = 28.3, SD = 9.0, range: [19, 66]; Sex: 48.3% females, 51.0% males, 0.7% others).

@@

The real-fake ratings (measured originally with a [-1, 1] analog scale) were converted into two scores, corresponding to two conceptually distinct mechanisms: the dichotomous *belief* (real or fake, based on the sign of the rating) and the *confidence* (the rating's absolute value) associated with that belief. The former was analyzed using logistic mixed models, which modelled the probability of assigning a face to the real (\>= 0) as opposed to fake (\< 0). The latter, as well as the other face ratings (attractiveness, beauty, trustworthiness and familiarity), was modelled using mixed beta regressions (suited for outcome variables expressed in percentages). The models included the participants and stimuli as random factors.

@@

<!-- ### Manipulation Check -->

@@

Several limitations have to be noted. The current experimental paradigm required participants to judge the realness of faces they had prior exposure to (which was done to prevent reality judgements from influencing the other ratings). Although the effect of re-exposure delay was negligible, the potential bias induced by face familiarity (as compared to judging completely new items) cannot be discarded. Future studies could examine that by incorporating novel face images or increasing the duration of the re-exposure delay.

A related issue is the possible contamination of the reality judgements by the prior explicit prompting

Moreover, the magnitude of the effects found in the study was relatively small, suggesting that the facial features measured in the study were not the key determinants of simulation monitoring. Hence, beyond exploring new potential mechanisms, future studies should include a more thorough debriefing to try to capture what conscious strategies (if any) the participants used (e.g., focusing on some features of the stimulus - like hair or eyes in the case of faces) to guide their reality beliefs.
